How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-02-22 doc/paper.tex
small paper changes

\begin{abstract}
  One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?

  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). Using a model of species survival, we analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future.

  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. Using \textit{k}-fold cross-validation, we found that in-sample model performance measures are approximately equal to out-of-sample performance, meaning that we can have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past. These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.

  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. The greatest source of variance in survival probability is the timing of observation. Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age.

  We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. However, the effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the globe) between any two occurrences; this measure is also called a geodesic. This distance was measured in kilometers. The geodesic distance was then log-plus-one transformed and then standardized by zero-centering the data and then dividing by its standard deviation so that geographic range had mean 0 and standard deviation 1. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, and that coefficients associated with similarly standardized covariates will be directly comparable in magnitude \citep{ARM}.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above).

@@

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\). These regression coefficients are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

This logistic regression model, minus the final selection of priors, is expressed as 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{d[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    % double check this\dots is A MVN dist?
    A_{w, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
  \end{aligned}
  \label{eq:core}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(-2, 5) & \text{if } d = \text{intercept} \\
        N(-1, 1) & \text{if } d = \text{geo. range} \\
        N(0, 1) & \text{else } \\
      \end{cases} \\
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
\begin{verbatim}
form <- event ~ range + range_diff + temp + temp_lag + 
                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                (1 | age/phylum), 
stan_glmer(formula = form,
           data = data, 
           family = 'binomial',
           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
           prior_intercept = normal(0, 10, autoscale = FALSE), 
           prior_aux = cauchy(0, 3, autoscale = FALSE), 
           chains = 4,
           thin = 4,
           adapt_delta = 0.999999)
\end{verbatim}

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ range + range_diff + temp + temp_lag +
           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
           (1 | age/phylum))
brmfit <- brm(formula = form,
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

@@

The results of this paper are a measure of baseline ability to predict relative differences in extinction risk. By focusing on two aspects of species macroecology -- geographic range and global temperature -- we developed a rather naive model for estimating the probability of a species extinction based on their occurrence records.

We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original). The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

The striking aspects of our in-sample results are the statistical differences in AUC between the models, and the incredible narrow band of estimated AUC values. For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible. This means that even though one model has a statistically greater AUC, the practical difference between the models is not scientifically significant.

This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses. By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored. For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the inverse-logit of the intercept is close to 0 or 1 CITATION.

The success of model is partially driven by the size of our dataset and the structure of our model. Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping. Instead of relying on large amounts of ecological information to shape individual differences, our model leverages the entire Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.

The reason we were not able to include more biological information in our models about the species studied is that we lack most any life history or ecological information on most marine micro- and nano-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species.

If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a subset of the fossil occurrences present the Neptune Database. This means ignoring the vast trove of occurrence information present in the rest of the database.

>>>>

\begin{abstract}
  One of the great promises of paleobiology is that by studying the past we can better predict the future.  This promise is particularly pertinent given as risk assessments for some modern species could potentially be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales.  Any effort to assess future risk based on past extinctions and range trajectories must address two key questions:  (1) At a given timescale, are geographic range and extinction risk trajectories deterministic (past trends are likely to continue into the future) or Markovian (the future depends only on the present state)? (2) Given knowledge of past extinction/survival patterns and the present geographic ranges of extant taxa, how accurate are extinction risk predictions?  
  To address these questions we analyze the fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We analyzed how survival probability changes over time as a function of species age, time of observation, current geographic range, most recent change in geographic range, global temperature average, and the lag of global temperature. %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  Our results show that our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. The greatest source of variance in survival probability is the timing of observation. Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

A taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this measure is also called a geodesic. This distance was measured in kilometers. Geographic range was then log-plus-one transformed, and standardized by zero-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that the analyzed values of geographic range had mean 0 and standard deviation 1. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the interecpt term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the transformed and standardized geographic range values and not standardized seperately.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above). Difference in temperature between between observations was measured from the transformed and standardized temperature values and not standardized seperately.

@@

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla. The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

The effect of species age on the log-odds of species extinction is modeled as a non-nested random intercept \(A\) \citep{Tutz2016}. This term describes how the log-odds of extinction varies along a species duration, and how this effect can differ between the phyla. \(A\) is a \(l\) by \(p\) matrix, where \(l\) is the age at observation of a species and \(p\) is its phylum. \(A\) is modeled as following a multivariate normal distribution with phylum means being vector \(\delta\) and covariance matrix \(\Sigma_{A}\). The covariation between the elements of vector \(\delta\) are modeled as a multivariate normal distribution with a mean vector of all 0s and covariance matrix \(\Sigma_{\delta}\).

@@

In total, our model can be expressed as: 
\begin{equation}
  \begin{aligned}
    t_{i} &\sim \text{Bernoulli}(\Theta) \\
    \Theta_{i} &= \text{logit}^{-1} (X_{i} B_{w[i], p[i]} + A_{l[i], p[i]}) \\
    B_{w, p} &\sim MVN(\alpha_{p}, \Sigma_{B}) \\
    \alpha_{p} &\sim MVN(\mu, \Sigma_{\alpha}) \\
    % double check this\dots is A MVN dist?
    A_{l, p} &\sim MVN(\delta_{p}, \Sigma_{A}) \\
    \delta_{p} &\sim \mathrm{N}(0, \sigma_{\delta}) \\
    \mu_{d} &\sim 
      \begin{cases}
        N(-2, 5) & \text{if } d = \text{intercept} \\
        N(-1, 1) & \text{if } d = \text{geo. range} \\
        N(0, 1) & \text{else } \\
      \end{cases} \\
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}
with \(i\) indexing the observation and bracket subscripts referencing the class of the \(i\)th observation where \(w[i]\) is the time of the \(i\)-th observation, \(p[i]\) is the phylum of the \(i\)-th observation, and \(d[i]\) is the age of the \(i\)-th observation.

@@

%An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
%\begin{verbatim}
%form <- event ~ range + range_diff + temp + temp_lag + 
%                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
%                (1 | age/phylum), 
%stan_glmer(formula = form,
%           data = data, 
%           family = 'binomial',
%           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
%           prior_intercept = normal(0, 10, autoscale = FALSE), 
%           prior_aux = cauchy(0, 3, autoscale = FALSE), 
%           chains = 4,
%           thin = 4,
%           adapt_delta = 0.999999)
%\end{verbatim}
%
%Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} \citep{brms2017,brms2018} as:
%\begin{verbatim}
%priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
%            set_prior('normal(0, 1)', class = 'b'),
%            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
%            set_prior('cauchy(0, 3)', class = 'sd'),
%            set_prior('lkj(1)', class = 'cor'))
%form <- bf(event ~ range + range_diff + temp + temp_lag +
%           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
%           (1 | age/phylum))
%brmfit <- brm(formula = form,
%              data = data, 
%              family = bernoulli(), 
%              prior = priors,
%              chains = 4, 
%              thin = 4,
%              control = list(adapt_delta = 0.999999)
%\end{verbatim}

@@

The results of this paper set out our baseline ability to predict relative differences in extinction risk. We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original). The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_ts}, \ref{fig:auc_taxon}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_time}, \ref{fig:fold_auc_taxon}).

Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though our ``past and vary'' model has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

This result is an important reminder about understanding the practical interpretation of our analyses, which can be lost when we do not consider the predictive aspects of our analyses. By focusing on determining which covariates are ``significant'' and which model is best through simple comparisons means that the practical importance of the results are ignored. For example, in logistic regression a covariate can be considered significant on the log-odds scale but have no practical difference on the probability scale because the range of values is too small as to matter such as when the intercept is greater than 2 or less than -2 because the inverse-logit of those values are close to 0 or 1, respectively \citep{ARM}.

The success of model is partially driven by the size of our dataset and the hierarchical structure of our model. Our estimates are based on rather limited information about the taxa themselves, and our model only takes into account some aspects of species geographic range and their rough taxonomic grouping. Instead of relying on large amounts of ecological information to shape individual differences, our model leverages most of the Neptune dataset through a multilevel model to constrain and improve our parameter estimates by sharing information about those parameters across taxa and time.

The principal reason we were not able to include more biological information in our models is that we lack most any life history or ecological information on most marine micro- and nano-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}. However, not this information does not exist all foram species. If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species. This means ignoring the majority of occurrence information present in the Neptune database.
